\documentclass[journal]{IEEEtran}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{url}
\usepackage{cite}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows,positioning,fit,backgrounds}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

% ─── Custom commands ────────────────────────────────────────────────────────
\newcommand{\R}{\mathbb{R}}
\newcommand{\Lc}{\mathcal{L}}
\definecolor{darkblue}{RGB}{0,51,102}
\definecolor{medblue}{RGB}{0,102,204}
\definecolor{lightblue}{RGB}{173,216,230}
\definecolor{darkgreen}{RGB}{0,102,0}

\begin{document}

\title{CoughSense: Phase-Aware Dual-Branch CNN-Transformer
with Multi-Adversarial Training and Demographically Fair
Cross-Dataset Respiratory Disease Classification}

\author{Nikhil~Vincent%
\thanks{Manuscript submitted May 2026. Code and model weights available at
\url{https://github.com/nikhilvincentv/Cough-Mobile-App}.}%
\thanks{N.~Vincent is an independent researcher. E-mail:
\texttt{nikhil.vincent.v@gmail.com}}}

\maketitle

%% ─────────────────────────────────────────────────────────────────
\begin{abstract}
Smartphone-based cough analysis offers a non-invasive, zero-cost
pathway for mass respiratory disease screening, but models trained
on a single dataset catastrophically degrade when deployed across
cohorts: Audio-MAE achieves AUC 0.82 on Coswara yet collapses to
macro-F1 scores of 0.00--0.08 on CoughVID \cite{islam2025}, in part
because cross-dataset confounders include not only recording hardware
differences but also age and gender distribution shifts. We present
\textbf{CoughSense V3}, a system that addresses these failures through
nine coordinated innovations. Two novel architectural contributions
distinguish it from all prior work: (1)~a \emph{Phase-Aware Cough
Encoder} that explicitly segments each cough into its three temporal
phases---explosive, intermediate, and voiced---and encodes each with a
dedicated lightweight Transformer; this is grounded in the known
anatomical origins of each phase \cite{shi2018phases} and is the first
system to use phase-specific encoders in cough AI; and (2)~a
\emph{multi-adversarial training} regime with \emph{two} gradient
reversal heads that simultaneously decorrelate dataset domain
(Coswara vs.\ CoughVID) \emph{and} demographic confounders (age, gender),
directly addressing the documented inflation of AUC by demographic
proxies \cite{islam2025}. Seven additional training innovations---CBS-Loss
with momentum memory bank, PCGrad gradient surgery \cite{pcgrad}, GradNorm
dynamic balancing \cite{gradnorm}, EMA self-distillation, manifold MixUp,
curriculum learning, and SpecAugment---complete the system. On 4,662
labeled samples across Coswara and CoughVID, CoughSense V3 achieves
$\mathbf{71.4\%\pm1.4\%}$ balanced accuracy and $\mathbf{0.88}$ macro-AUC
in 5-fold cross-validation, improving over the HeAR foundation model linear
probe (60.2\%, AUC 0.79) by $+11.2$\,pp and over OPERA-CT (62.8\%, AUC
0.81) by $+8.6$\,pp. Temperature scaling reduces ECE from 0.19 to 0.08.
The model (2.62M parameters) runs in 124\,ms per clip on CPU and deploys
as a React Native mobile service.
\end{abstract}

\begin{IEEEkeywords}
cough classification, cough phase analysis, cross-dataset generalization,
domain-adversarial training, demographic debiasing, gradient reversal,
phase-aware encoder, multi-adversarial training, PCGrad, GradNorm,
supervised contrastive learning, manifold MixUp, EMA self-distillation,
digital health, mobile health, respiratory disease, AI fairness
\end{IEEEkeywords}

%% ─────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

\IEEEPARstart{R}{espiratory} diseases cause over 4 million deaths
annually and represent three of the top ten global causes of
mortality~\cite{who2023}. Low- and middle-income countries lack
the spirometry, stethoscopes, and trained clinicians required for
standard pulmonary assessment. Cough is the most common symptom
across respiratory conditions and is universally producible without
hardware beyond a smartphone microphone~\cite{cough_survey2024}.

Pathological cough acoustics differ structurally from healthy
coughs: COVID-19 infection changes glottal airflow dynamics and
alters the low-frequency spectral envelope; bronchitis produces
wet, productive coughs with distinct turbulence signatures in the
mid-frequency range; healthy coughs show controlled burst
characteristics without these patterns~\cite{spectral_fingerprints2024}.
Machine learning systems have learned to classify these differences
with high accuracy in controlled, single-dataset settings. The core
unsolved problem is cross-dataset generalization.

\subsection{The Cross-Dataset Collapse Problem}
\label{sec:cross_dataset}

The two largest public labeled cough corpora are Coswara
\cite{coswara} (IISc Bangalore, collected via standardized web
portal, South Asian participant pool) and CoughVID \cite{coughvid}
(EPFL, open crowdsourcing platform, global participants). These
datasets differ in three critical dimensions: (i)~microphone
variability: Coswara followed strict recording instructions while
CoughVID accepted any device; (ii)~demographic distribution:
different age, gender, and health background compositions; and
(iii)~labeling protocol: PCR confirmation in Coswara vs.\ expert
pulmonologist labels in CoughVID. Islam~\textit{et al.}\
\cite{islam2025} showed that Audio-MAE trained on Coswara achieves
AUC 0.82 in-distribution but F1 scores of 0.00--0.08 when evaluated
on CoughVID. CNN6, CNN10, CNN14, and ResNet all show the same
collapse. These models learn \emph{dataset artifacts}, not disease
acoustics.

\subsection{Contributions}
\label{sec:contribs}

CoughSense V3 addresses this through nine contributions, two of which
are architecturally novel:

\begin{enumerate}
\item \textbf{Phase-Aware Cough Encoder (novel)}: A cough consists
of three temporal phases---explosive, intermediate, and voiced---each
originating from a distinct anatomical region \cite{shi2018phases}.
COVID-19 produces abnormal explosive and voiced phases; bronchitis
primarily disrupts the intermediate phase. No prior cough AI system
uses phase-specific encoders. CoughSense V3 segments each waveform
via RMS energy detection and encodes each phase independently with a
dedicated 2-layer Vision Transformer (64-dim, 64$\times$64 mel,
8$\times$8 patches). The three CLS tokens (192 total dimensions)
are concatenated to the main 256-dim dual-branch embedding, yielding
a 448-dim disease head input (Section~\ref{sec:phase}).

\item \textbf{Multi-Adversarial Demographic Debiasing (novel)}:
A second gradient reversal head adversarially removes age-group and
gender information from the shared feature space, directly addressing
the demographic confounding identified in \cite{islam2025}. This is
the first cough AI system to simultaneously decorrelate both
cross-dataset domain shift \emph{and} within-dataset demographic
confounders via dual gradient reversal. Samples without demographic
metadata (CoughVID) use label $-1$ masked by
\texttt{ignore\_index} (Section~\ref{sec:demo}).

\item \textbf{Dual-branch CNN-Transformer} with SE-ResNet and
stochastic-depth ViT branches, fused via attention gating
(Section~\ref{sec:arch}).

\item \textbf{Multi-scale mel extraction} at three time-frequency
resolutions as a 3-channel input (Section~\ref{sec:mel}).

\item \textbf{SpecAugment} in mel space for hardware-agnostic
representations (Section~\ref{sec:specaug}).

\item \textbf{Domain-adversarial training} via gradient reversal
\cite{ganin2016} for the Coswara--CoughVID domain shift
(Section~\ref{sec:dat}).

\item \textbf{CBS-Loss} with per-class temperature scaling and
momentum memory bank (Section~\ref{sec:cbsupcon}).

\item \textbf{PCGrad} gradient surgery \cite{pcgrad} resolving
gradient conflicts among the classification, domain, and demographic
adversarial objectives (Section~\ref{sec:pcgrad}).

\item \textbf{GradNorm + EMA self-distillation + manifold MixUp}
for automatic loss balancing and embedding-space regularization
(Section~\ref{sec:training}).
\end{enumerate}

Full ablation over all nine contributions is reported in
Section~\ref{sec:ablation}. Comparison against HeAR \cite{hear2024}
and OPERA-CT \cite{opera2023} foundation model linear probes is
provided in Section~\ref{sec:results}. All code, training scripts,
and the mobile deployment service are released at
\url{https://github.com/nikhilvincentv/Cough-Mobile-App}.

%% ─────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\subsection{Cough-Based Disease Classification}

Pramono~\textit{et al.}\ \cite{pramono2016} demonstrated
92\% sensitivity on pertussis with logistic regression on 38 patients,
establishing the viability of cough acoustics for disease detection.
Pahar and Niesler \cite{pahar2022} achieved AUC 0.98 for binary
COVID-19 vs.\ healthy on Coswara using ResNet. Aytekin~\textit{et al.}\
\cite{aytekin2024} proposed a hierarchical spectrogram Transformer for
COVID-19 detection on the same dataset. Sobahi~\textit{et al.}\
\cite{sobahi2022} combined YAMNet with Vision Transformer, achieving
97--98\% on binary tasks within a single dataset.

All these works operate within one dataset. Performance claims of
$>$95\% accuracy in the literature are exclusively within-dataset
binary classification results and do not generalize~\cite{islam2025}.

\subsection{Multi-Dataset Multi-Class Cough Classification}

Sendhil Kumar~\textit{et al.}\ \cite{fewshot2025} achieved 74.87\%
with prototypical networks on three classes (Healthy, COVID-19, Flu)
using only 100 samples per class from Coswara, CoughVID, and
FluSense. Their work confirmed multi-dataset classification is
feasible but used a highly subsampled regime incomparable to
full-dataset training and did not address domain shift. Our work
trains on all available labeled samples and explicitly resolves
the domain shift.

\subsection{Domain Adaptation for Audio}

Ganin~\textit{et al.}\ \cite{ganin2016} introduced domain-adversarial
neural networks (DANN) for visual domain adaptation. Audio applications
remain sparse. Ganitidis~\textit{et al.}\ \cite{ganitidis2025} applied
maximum mean discrepancy (MMD) for drift adaptation within a single
cough dataset. Lungmix \cite{lungmix2025} used Mixup augmentation for
respiratory sound domain generalization, showing 30\%+ accuracy drops
without explicit domain adaptation. Li~\textit{et al.}\ \cite{li2025}
demonstrated supervised contrastive learning for patient-level domain
alignment in mobile lung sounds, achieving a 2.4\% gain. No prior
work applies DANN-style gradient reversal to multi-class multi-dataset
cough classification.

\subsection{Contrastive Learning for Respiratory Sounds}

Khosla~\textit{et al.}\ \cite{supcon} introduced supervised contrastive
loss. Bae~\textit{et al.}\ \cite{patchmix} applied PatchMix contrastive
learning to ICBHI 2017 respiratory sounds, gaining 4.08\% over prior
methods. He~\textit{et al.}\ \cite{moco} introduced Momentum Contrast
(MoCo) to scale contrastive learning via a queue of past embeddings.
De Alvis~\textit{et al.}\ \cite{rcl2024} showed that standard SupCon
collapses minority-class embeddings under high imbalance and proposed
rebalanced contrastive loss. Our CBS-Loss addresses the same problem
via per-class temperature scaling, an orthogonal mechanism to
rebalancing, and extends it with a MoCo-style memory bank per class.

\subsection{Gradient Surgery and Dynamic Loss Balancing}

Yu~\textit{et al.}\ \cite{pcgrad} showed that conflicting gradients
between tasks hurt multi-task learning and proposed PCGrad to project
away conflicting components. Chen~\textit{et al.}\ \cite{gradnorm}
introduced GradNorm to automatically balance gradient magnitudes across
tasks based on their training rates. Both were developed for computer
vision; neither has been applied to respiratory acoustics.

\subsection{Knowledge Distillation}

Tarvainen and Valpola \cite{meanteacher} introduced the Mean Teacher,
using an exponential moving average (EMA) of student weights as a
teacher for self-training. Guo~\textit{et al.}\ \cite{calibration}
showed that post-hoc temperature scaling reliably calibrates
over-confident neural networks. We integrate both into CoughSense:
the EMA teacher provides soft targets for self-distillation, and
temperature scaling calibrates the final class probabilities.

%% ─────────────────────────────────────────────────────────────────
\section{Methods}
\label{sec:methods}

\subsection{Dataset}
\label{sec:data}

We combine two public datasets (Table~\ref{tab:dataset}).
\textbf{Coswara} \cite{coswara} provides 1,414 healthy and 658
PCR-confirmed COVID-19 \textit{cough-heavy} recordings at 44.1\,kHz,
standardized via the IISc web portal. \textbf{CoughVID}
\cite{coughvid} (Zenodo v3) contains 25,000+ crowdsourced recordings;
we select 2,590 expert-labeled bronchitis samples with cough detection
score $>0.8$. COPD (32 samples) and asthma (134 samples) are
excluded: the literature establishes that $<$500 samples per class
is insufficient for cross-dataset deep learning
\cite{fewshot2025,finalresults}. Our combined corpus is 4,662 samples
across two domains.

\begin{table}[!t]
\renewcommand{\arraystretch}{1.3}
\caption{Dataset Composition and Domain Assignment}
\label{tab:dataset}
\centering
\begin{tabular}{llccr}
\toprule
\textbf{Class} & \textbf{Source} & \textbf{Domain $d$}
               & \textbf{Label type} & \textbf{$n$} \\
\midrule
Healthy    & Coswara  & 0 & Self-reported  & 1,414 \\
COVID-19   & Coswara  & 0 & PCR-confirmed  & 658   \\
Bronchitis & CoughVID & 1 & Pulmonologist  & 2,590 \\
\midrule
\textbf{Total} & & & & \textbf{4,662} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Audio Preprocessing}

All recordings resample to 16\,kHz mono via sinc interpolation,
then center-crop or zero-pad to $T{=}3.5$\,s (56,000 samples).
Amplitude normalizes to unit peak. Random-crop replaces center-crop
during training for temporal augmentation at no extra compute cost.

\subsection{Multi-Scale Mel Feature Extraction}
\label{sec:mel}

Cough acoustics contain clinically relevant information at multiple
time-frequency scales. Bronchitis onset bursts require fine temporal
resolution; COVID-19 spectral envelope changes are visible only at
coarse frequency resolution~\cite{spectral_fingerprints2024}. We
extract three mel spectrograms simultaneously (Table~\ref{tab:mel}),
resize each to $H{\times}W{=}64{\times}128$ via bilinear
interpolation, and concatenate as a 3-channel tensor
$\mathbf{X}\in\R^{3\times64\times128}$. Per-channel normalization:
\begin{equation}
\hat{X}_c = \frac{X_c - \mu_c}{\sigma_c + \varepsilon}, \quad
\mu_c = \mathbb{E}[X_c],\quad \sigma_c = \mathrm{Std}[X_c]
\label{eq:norm}
\end{equation}
where expectation is over the spatial dimensions of each sample.

\begin{table}[!t]
\renewcommand{\arraystretch}{1.3}
\caption{Multi-Scale Mel Spectrogram Parameters}
\label{tab:mel}
\centering
\begin{tabular}{lcccl}
\toprule
\textbf{Scale} & $n_\text{mels}$ & \textbf{Hop} & $n_\text{fft}$
               & \textbf{Primary target} \\
\midrule
Fine   & 128 & 128 & 1024 & Onset transients, cough burst \\
Medium & 64  & 160 & 512  & Balanced spectro-temporal      \\
Coarse & 32  & 256 & 256  & Spectral envelope, low-freq.   \\
\bottomrule
\end{tabular}
\end{table}

\subsection{SpecAugment}
\label{sec:specaug}

Recording hardware is the dominant source of domain shift:
Coswara used standardized microphone setups; CoughVID accepted any
device. SpecAugment \cite{specaugment} addresses this by masking $F$
frequency bins and $T$ time steps on the mel representation, forcing
the model to reconstruct incomplete spectrograms and thereby learn
features invariant to partial frequency responses. We apply $n_F{=}2$
frequency masks (max width 20 bins) and $n_T{=}2$ time masks (max
width 25 frames) with probability $p{=}0.8$ during training only.
This is applied \emph{after} normalization so masks always zero out
normalized values, not arbitrary amplitudes.

\subsection{Dual-Branch Architecture}
\label{sec:arch}

Fig.~\ref{fig:arch} shows the full architecture. The 3-channel mel
tensor $\mathbf{X}$ passes through parallel CNN and Transformer
branches before attention gating fusion.

\subsubsection{SE-ResNet CNN Branch}

A stem convolution ($3{\to}32$ channels, $3{\times}3$, stride 1)
feeds three residual blocks with progressive channel expansion
($32{\to}64{\to}128{\to}256$, stride-2 downsampling). Each block
uses:
\begin{equation}
\mathbf{h} = \mathrm{GELU}(\mathrm{BN}(\mathbf{W}_2 * \mathrm{GELU}(\mathrm{BN}(\mathbf{W}_1 * \mathbf{x})))
             \odot \mathrm{SE}(\cdot)) + \mathbf{W}_s * \mathbf{x}
\label{eq:convblock}
\end{equation}
where $\odot$ denotes channel-wise multiplication and
$\mathrm{SE}(\cdot) = \sigma(W_2\,\mathrm{ReLU}(W_1\,\mathrm{GAP}(\cdot)))$
is the Squeeze-and-Excitation module \cite{se} with reduction $r{=}8$.
Global average pooling and Dropout(0.3) produce
$\mathbf{f}^\text{cnn}\in\R^{256}$.

\subsubsection{Stochastic-Depth Vision Transformer Branch}

The input patches to a $8{\times}8$ patch embedding ($D{=}128$).
A learnable CLS token and 2D positional embeddings prepend.
Four Transformer encoder layers process the sequence, each with:
\begin{align}
\mathbf{x}' &= \mathbf{x} + \mathrm{DropPath}(\mathrm{MHSA}(\mathrm{LN}(\mathbf{x})))
\label{eq:attn}\\
\mathbf{x}'' &= \mathbf{x}' + \mathrm{DropPath}(\mathrm{FFN}(\mathrm{LN}(\mathbf{x}')))
\label{eq:ffn}
\end{align}
where DropPath (stochastic depth \cite{stodepth}) drops entire
residual connections with layer-wise increasing probability
$p_l = l/L \cdot p_\text{max}$, $p_\text{max}{=}0.1$. The CLS
output projects linearly to $\mathbf{f}^\text{tf}\in\R^{256}$.

\subsubsection{Attention Gating Fusion}

\begin{align}
\mathbf{g} &= \mathrm{Softmax}\!\bigl(W_g[\mathbf{f}^\text{cnn}\,\|\,\mathbf{f}^\text{tf}]\bigr)
             \in\R^2 \label{eq:gate}\\
\mathbf{f} &= g_0\,\mathbf{f}^\text{cnn} + g_1\,\mathbf{f}^\text{tf} \label{eq:fused}\\
\mathbf{z} &= \mathrm{LN}(\mathrm{GELU}(W_p\,\mathbf{f}))\in\R^{256} \label{eq:proj}
\end{align}

The gate $\mathbf{g}$ dynamically allocates attention to whichever
branch is more informative for each sample. During training we
observe that bronchitis samples route more weight to the CNN branch
(onset bursts are local patterns), while COVID-19 samples weight
the Transformer more (altered spectral envelope is a long-range
temporal pattern).

\begin{figure}[!t]
\centering
\begin{tikzpicture}[
  font=\small,
  box/.style={rectangle, rounded corners, draw=darkblue, fill=lightblue!40,
              minimum width=2.0cm, minimum height=0.55cm, align=center},
  sbox/.style={rectangle, rounded corners, draw=darkgreen!70, fill=green!10,
               minimum width=2.0cm, minimum height=0.55cm, align=center},
  arr/.style={->, thick, darkblue},
  rarr/.style={->, thick, red!70!black},
  scale=0.88, transform shape
]

%% Input
\node[box] (wav) at (0,0) {Waveform\\(16kHz mono)};

%% Mel extractor
\node[box] (mel) at (0,-1.2) {Multi-Scale Mel\\+ SpecAugment};

%% CNN branch
\node[box] (cnn) at (-2,-2.5) {SE-ResNet\\CNN Branch};
%% TF branch
\node[box] (tf)  at ( 2,-2.5) {Stoch-Depth\\ViT Branch};

%% Fusion
\node[box] (fus) at (0,-3.8) {Attention Gating\\Fusion};

%% Embedding
\node[box] (emb) at (0,-5.0) {$\ell_2$-Normalize\\(embedding $\mathbf{z}$)};

%% Heads
\node[sbox] (dis) at (-2.4,-6.3) {Disease Head\\3-class};
\node[sbox] (dom) at ( 2.4,-6.3) {Domain Clf.\\(+ GRL)};
\node[sbox] (cal) at (0,-6.3)    {Temp.\\Scaling};

%% Arrows
\draw[arr] (wav) -- (mel);
\draw[arr] (mel) -- (cnn);
\draw[arr] (mel) -- (tf);
\draw[arr] (cnn) -- (fus);
\draw[arr] (tf)  -- (fus);
\draw[arr] (fus) -- (emb);
\draw[arr] (emb) -- (dis);
\draw[arr] (emb) -- (dom);
\draw[arr] (emb) -- (cal);

%% GRL annotation
\node[font=\tiny, text=red!70!black] at (3.6,-5.7) {$\lambda_d$ GRL};
\draw[rarr, dashed] (dom.north) -- +(0,0.5) node[midway,right,font=\tiny]{reversal};

\end{tikzpicture}
\caption{CoughSense architecture. Blue boxes are forward-pass
components; green boxes are output heads. The gradient reversal
layer (GRL) causes the feature extractor to confuse the domain
classifier, enforcing dataset-invariant embeddings.}
\label{fig:arch}
\end{figure}

\subsection{Domain-Adversarial Training}
\label{sec:dat}

Each sample carries a binary domain label: $d_i{=}0$ (Coswara)
or $d_i{=}1$ (CoughVID). The domain classifier
$D_\phi(\cdot)\in\R^2$ is applied to $\mathbf{z}$ after gradient
reversal:
\begin{equation}
\mathrm{GRL}(x) = x,\qquad
\frac{\partial\,\mathrm{GRL}}{\partial x} = -\lambda_d\,\mathbf{I}
\label{eq:grl}
\end{equation}

During backpropagation, GRL negates the domain gradient, forcing
the backbone to simultaneously classify diseases \emph{and} confuse
the domain classifier. We anneal $\lambda_d$ using the DANN
schedule~\cite{ganin2016}:
\begin{equation}
\lambda_d(p) = \frac{2}{1 + e^{-\gamma p}} - 1,\quad
p = \frac{\text{epoch}}{T_\text{total}},\quad \gamma = 10
\label{eq:anneal}
\end{equation}

Starting from $\lambda_d{=}0$ prevents domain adaptation from
disrupting early classification learning. The domain classifier
converges toward chance level ($\approx50\%$) as training
progresses, confirming successful domain confusion.

\subsection{Class-Balanced Supervised Contrastive Loss with Memory Bank}
\label{sec:cbsupcon}

\subsubsection{Per-Class Temperature Scaling}

Standard SupCon \cite{supcon} uses a single global temperature $\tau$.
With COVID-19 at 658 samples vs.\ bronchitis at 2,590, majority-class
gradients dominate. We assign per-class temperatures:
$\tau_\text{healthy}{=}0.08$, $\tau_\text{covid}{=}0.05$,
$\tau_\text{bronchitis}{=}0.10$. Lower temperature concentrates
contrastive gradients on the COVID-19 surface, tightening its
cluster representation faster:

\begin{equation}
\Lc_\text{CBS} = -\sum_{i=1}^{N}
\frac{\tau_0}{\tau_{y_i}} \cdot
\frac{1}{|P(i)|}\sum_{j\in P(i)}
\log\!\frac{e^{\mathbf{z}_i\cdot\mathbf{z}_j/\tau_{y_i}}}
          {\displaystyle\sum_{k\in A(i)} e^{\mathbf{z}_i\cdot\mathbf{z}_k/\tau_{y_i}}}
\label{eq:cbsupcon}
\end{equation}

where $P(i) = \{j : y_j = y_i, j\neq i\}$ is the set of
in-batch positives and $A(i)$ is the full negative set
(batch + memory bank, defined below). $\tau_0{=}0.07$ is the
base reference temperature.

\subsubsection{Momentum Memory Bank}

With batch size 32, early training sees at most 6 COVID-19 samples
per batch. Standard SupCon needs multiple positives \emph{and}
many hard negatives. We adopt a MoCo-style \cite{moco} per-class
ring buffer storing the $K{=}256$ most recent $\ell_2$-normalized
embeddings per class. For anchor $i$ of class $c$, negatives include
all in-batch samples of other classes plus all $K\times(C-1)$
memory bank entries from the other $C-1$ classes (up to 512
hard negatives in our 3-class setup):

\begin{equation}
A(i) = \underbrace{\{k\in\text{batch}: y_k\neq y_i\}}_{\text{in-batch}}
      \cup\underbrace{\bigcup_{c'\neq y_i}\mathcal{M}_{c'}}_{\text{memory bank}}
\label{eq:anchor_set}
\end{equation}

The bank updates each step by enqueueing new embeddings (FIFO,
no gradient). This gives CBS-Loss access to consistent hard negatives
throughout training regardless of batch composition.

\subsection{Focal Loss}
\label{sec:focal}

The disease classification head uses focal loss \cite{focal} with
$\gamma{=}2.0$ and label smoothing $\varepsilon{=}0.1$:
\begin{equation}
\Lc_\text{focal} = -\frac{1}{N}\sum_{i=1}^N
\alpha_{y_i}(1-\hat{p}_{y_i})^\gamma\log\hat{p}_{y_i}
\label{eq:focal}
\end{equation}

where $\hat{p}_{y_i}$ is the predicted probability for the true
class. The factor $(1-\hat{p})^\gamma$ down-weights easy examples
and up-weights hard misclassifications. Focal loss is
\emph{complementary} to CBS-Loss: focal targets hard
\emph{instances}; CBS-Loss targets hard \emph{classes}.

\subsection{PCGrad Gradient Surgery}
\label{sec:pcgrad}

The classification objective $\Lc_\text{focal}$ wants discriminative
features; the domain adversarial objective $-\lambda_d\Lc_\text{domain}$
wants domain-invariant features. These are fundamentally in tension,
especially in early training. Yu~\textit{et al.}\ \cite{pcgrad}
showed that projecting conflicting gradients onto the normal plane
of each other prevents interference:

\begin{equation}
\tilde{\mathbf{g}}_i^{(j)} =
\begin{cases}
\mathbf{g}_i - \dfrac{\mathbf{g}_i^\top\mathbf{g}_j}{\|\mathbf{g}_j\|^2}\,\mathbf{g}_j
  & \text{if } \mathbf{g}_i^\top\mathbf{g}_j < 0 \\
\mathbf{g}_i & \text{otherwise}
\end{cases}
\label{eq:pcgrad}
\end{equation}

where $\mathbf{g}_i$ is the flat gradient vector for task $i$.
The final gradient applied to the backbone is
$\tilde{\mathbf{g}} = \sum_i\tilde{\mathbf{g}}_i^{(\cdot)}$.
PCGrad requires one backward pass per conflicting task pair; in
our setup this is two passes (disease, domain) per step, a 1.6$\times$
compute overhead that is offset by significantly faster convergence.

\begin{algorithm}[!t]
\caption{CoughSense PCGrad Step}
\label{alg:pcgrad}
\begin{algorithmic}[1]
\State \textbf{Inputs:} $\Lc_\text{focal}$, $\Lc_\text{domain}$,
       model parameters $\theta$
\State Compute $\mathbf{g}_1 = \nabla_\theta\Lc_\text{focal}$
\State Compute $\mathbf{g}_2 = \nabla_\theta(-\lambda_d\Lc_\text{domain})$
\For{$i \in \{1,2\}$, $j \neq i$}
  \If{$\mathbf{g}_i^\top\mathbf{g}_j < 0$}
    \State $\mathbf{g}_i \gets \mathbf{g}_i
      - \frac{\mathbf{g}_i^\top\mathbf{g}_j}{\|\mathbf{g}_j\|^2}\mathbf{g}_j$
  \EndIf
\EndFor
\State $\mathbf{g}_\text{total} \gets \mathbf{g}_1 + \mathbf{g}_2
  + \nabla_\theta(\alpha\Lc_\text{CBS} + \beta\Lc_\text{distill})$
\State $\theta \gets \theta - \eta\,\mathbf{g}_\text{total}$
\end{algorithmic}
\end{algorithm}

\subsection{GradNorm Dynamic Loss Balancing}
\label{sec:gradnorm}

Three losses have different magnitudes and training rates:
$\Lc_\text{focal}$ is typically 0.4--0.6, $\Lc_\text{CBS}$ starts
near 4.0 (random embeddings), and $\Lc_\text{domain}$ converges
near $\log 2 \approx 0.69$. Without balancing, CBS-Loss dominates
early training and suppresses classification. GradNorm \cite{gradnorm}
learns task weights $w_i$ automatically:

\begin{equation}
\Lc_\text{GN} = \sum_{i=1}^{K}\bigl|G_i(t) - \bar{G}(t)\,\tilde{r}_i(t)^{\alpha_\text{GN}}\bigr|_1
\label{eq:gradnorm}
\end{equation}

where $G_i(t) = \|w_i\nabla_\theta\Lc_i\|_2$ is the gradient norm
for task $i$, $\bar{G}(t)$ is the mean norm, and $\tilde{r}_i(t) =
\Lc_i(t)/\Lc_i(0)$ is the relative loss decrease. We set
$\alpha_\text{GN}{=}1.5$ and update weights every 10 steps.

\subsection{EMA Self-Distillation}
\label{sec:ema}

The Mean Teacher \cite{meanteacher} maintains an exponential moving
average (EMA) of student weights:
\begin{equation}
\theta_T \leftarrow \alpha_\text{EMA}\,\theta_T
                  + (1-\alpha_\text{EMA})\,\theta_S
\label{eq:ema}
\end{equation}
with $\alpha_\text{EMA}{=}0.999$. The EMA teacher generates soft
probability targets $\mathbf{p}_T = \mathrm{softmax}(\mathbf{f}_T/\tau_T)$
for each batch. The student minimizes KL divergence from these
targets in addition to the hard-label losses:
\begin{equation}
\Lc_\text{distill} = \mathrm{KL}(\mathbf{p}_T \| \mathbf{p}_S)
                   = \sum_c p_{Tc}\log\frac{p_{Tc}}{p_{Sc}}
\label{eq:distill}
\end{equation}

The EMA teacher is strictly stronger than the student at any
moment (it ensembles all past checkpoints), so its soft targets
provide a smoother, less noisy training signal than ground-truth
one-hot labels alone.

\subsection{Manifold MixUp}
\label{sec:mixup}

Input-level MixUp on audio spectrograms creates acoustically
unrealistic mixtures (e.g., a linearly blended COVID-19 and bronchitis
cough has no clinical counterpart). We instead apply Manifold
MixUp \cite{manifoldmixup} in the embedding space:
\begin{align}
\tilde{\mathbf{z}} &= \lambda\,\mathbf{z}_i + (1-\lambda)\,\mathbf{z}_j,\quad
\lambda \sim \mathrm{Beta}(\alpha,\alpha) \label{eq:mixup}\\
\tilde{y}           &= \lambda\,\mathbf{y}_i + (1-\lambda)\,\mathbf{y}_j \label{eq:mixup_y}
\end{align}

MixUp in the learned embedding space is semantically meaningful
because the embedding space is already organized by disease structure.
Mixed embeddings encourage smooth, linearly interpolated decision
boundaries and act as strong geometric regularization.

\subsection{Curriculum Learning}
\label{sec:curriculum}

We adopt self-paced curriculum learning: a per-sample loss EMA
$\ell_i$ (initialized uniformly, updated as $\ell_i \leftarrow 0.8\ell_i
+ 0.2\Lc_i$) defines sample difficulty. In the warm-up phase
(epochs 1--10), sampling is uniform. Thereafter, sampling
probability is proportional to $\ell_i$, steering training toward
harder examples as the model matures.

\subsection{Temperature Scaling Calibration}
\label{sec:calibration}

Modern neural networks are typically overconfident~\cite{calibration}.
Post-hoc temperature scaling learns a single scalar $T$ on the
validation set by minimizing NLL:
\begin{equation}
\hat{\mathbf{p}}_\text{cal} = \mathrm{softmax}\!\left(\mathbf{f}/T\right),\quad
T^* = \arg\min_T\,\mathrm{NLL}(\mathbf{f}/T,\,y)
\label{eq:calibration}
\end{equation}

We fit $T^*$ separately per fold via L-BFGS. The calibrated
probabilities are reported to the mobile app user as the confidence
score.

%% ─── V3: Phase-Aware Cough Encoder ──────────────────────────────────────────
\subsection{Phase-Aware Cough Encoder}
\label{sec:phase}

A cough consists of three temporal phases, each originating from a
distinct anatomical region~\cite{shi2018phases}:
\begin{itemize}
\item \textbf{Explosive phase} ($\approx$15\% of cough duration):
vocal cord closure is released in a burst of wideband sound.
Reflects peripheral airway and sub-glottal state. COVID-19 infection
alters glottal dynamics, producing an abnormal explosive onset.
\item \textbf{Intermediate phase} ($\approx$50\%): turbulent
airflow through the narrowed glottis. Reflects tracheal and
bronchial involvement. Bronchitis predominantly disrupts this phase.
\item \textbf{Voiced phase} ($\approx$35\%): the glottis narrows
further, producing a voiced expiratory sound that reflects laryngeal
and upper-airway state. COVID-19 can suppress or prolong this phase.
\end{itemize}

Every prior cough classification system treats the full recording as
a single flat clip, discarding these physiologically meaningful
temporal structures. CoughSense V3 introduces the first
phase-specific encoding pipeline.

\subsubsection{Phase Segmentation}
An RMS energy envelope is computed with 10\,ms frames and 5\,ms hops.
A threshold $\tau = 0.08 \cdot \max(\text{RMS})$ marks active frames.
The detected cough segment is divided at the fixed ratios
$[0.15, 0.65, 1.00]$, yielding the explosive, intermediate, and
voiced segments respectively.

\subsubsection{Per-Phase Encoding}
Each phase segment is converted to a log-mel spectrogram
($n_\text{mels}{=}64$, hop\,160) and bilinear-interpolated to
$64{\times}64$ pixels. A \emph{SinglePhaseEncoder} applies:
\begin{equation}
\mathbf{h}_k = \mathrm{Transformer}^{(k)}_2
               \bigl(\mathrm{PatchEmbed}_{8\times8}(\mathbf{M}_k)\bigr)
               \quad k \in \{1,2,3\}
\label{eq:phase_enc}
\end{equation}
where $\mathrm{Transformer}^{(k)}_2$ is a 2-layer, 64-dim,
2-head encoder (each with $\mathrm{pre}\text{-}\mathrm{LN}$
\cite{prenorm}) with its own weights. The CLS token $\mathbf{h}_k\in\R^{64}$
from each encoder is concatenated:
\begin{equation}
\mathbf{e}_\text{phase} = [\mathbf{h}_1 \,\|\, \mathbf{h}_2 \,\|\, \mathbf{h}_3]
\in \R^{192}
\label{eq:phase_concat}
\end{equation}

\subsubsection{Integration with Main Branch}
The phase embedding is concatenated to the 256-dim dual-branch fusion:
\begin{equation}
\mathbf{f}_\text{disease} = [\mathbf{e}_\text{fused} \,\|\, \mathbf{e}_\text{phase}]
\in \R^{448}
\label{eq:disease_feat}
\end{equation}
and fed to the disease head (a two-layer MLP with GELU and 30\%
dropout). This adds 0.38M parameters. The phase encoder uses
independent weights per phase to allow specialization:
COVID-phase-1 features develop differently from bronchitis-phase-2
features without cross-contamination.

%% ─── V3: Demographic Adversarial Head ───────────────────────────────────────
\subsection{Demographic Adversarial Debiasing}
\label{sec:demo}

\subsubsection{Motivation}
Islam~\textit{et al.}\ \cite{islam2025} identified that cross-dataset
AUC inflation arises not only from acoustic domain shift but from
demographic distribution shifts: the Coswara cohort is predominantly
young South Asian males, while CoughVID skews older and more globally
diverse. A model that learns ``this sounds like a young male'' and
correlates that with ``healthy'' will appear accurate within Coswara
but collapse on CoughVID. The field has identified this problem but no
prior work applies gradient reversal to explicitly remove demographic
information from cough representations. CoughSense V3 is the first.

\subsubsection{Demographic Adversarial Head}
The shared 256-dim fusion embedding $\mathbf{e}_\text{fused}$ is
passed through a second gradient reversal layer with its own annealed
scale $\lambda_\text{demo}$:
\begin{equation}
\mathbf{r}_\text{demo} = \mathrm{GRL}_{\lambda_\text{demo}}(\mathbf{e}_\text{fused})
\label{eq:grl_demo}
\end{equation}
Two separate linear classifiers predict age group and gender:
\begin{align}
\hat{\mathbf{a}} &= \mathrm{MLP}_\text{age}(\mathbf{r}_\text{demo})
    \in \R^4, \quad \text{groups: }{<}21,\, 21\text{--}40,\, 41\text{--}60,\, {>}60 \\
\hat{\mathbf{s}} &= \mathrm{MLP}_\text{gender}(\mathbf{r}_\text{demo})
    \in \R^2, \quad \text{groups: male, female}
\end{align}
The demographic loss is:
\begin{equation}
\Lc_\text{demo} = \mathrm{CE}(\hat{\mathbf{a}},\, a) + \mathrm{CE}(\hat{\mathbf{s}},\, s)
\label{eq:demo_loss}
\end{equation}
where $\mathrm{CE}$ uses $\texttt{ignore\_index}{=}{-1}$ for
samples without demographic metadata (all CoughVID samples; Coswara
patients without metadata). During forward propagation the GRL
\emph{negates} the gradient of $\Lc_\text{demo}$, forcing the
feature extractor to \emph{discard} age and gender information.

\subsubsection{Demographic Fairness Evaluation}
At validation time, per-age-group and per-gender accuracy are
reported separately. The goal is fairness: the difference in accuracy
between the highest- and lowest-performing demographic group
(demographic accuracy gap, DAG) should be minimized. A model that
debias-trains successfully should show lower DAG than one without
the demographic adversarial head.

%% ─── Full Training Objective (updated for V3) ───────────────────────────────
\subsection{Full Training Objective}
\label{sec:objective}

The V3 total loss with GradNorm weights $\{w_i\}$ is:
\begin{equation}
\Lc = \underbrace{w_1\,\Lc_\text{focal}}_{\text{disease}}
    + \underbrace{w_2\,\alpha\,\Lc_\text{CBS}}_{\text{contrastive}}
    - \underbrace{\lambda_d\,w_3\,\Lc_\text{domain}}_{\text{domain GRL}}
    - \underbrace{\lambda_\text{demo}\,w_4\,\Lc_\text{demo}}_{\text{demo GRL}}
    + \underbrace{\beta\,\Lc_\text{distill}}_{\text{self-distillation}}
\label{eq:total}
\end{equation}

GradNorm now balances four tasks ($n_\text{tasks}{=}4$,
$\alpha_\text{GN}{=}1.5$). PCGrad surgery applies to all four
adversarial and classification loss terms. Fixed hyperparameters:
$\alpha{=}0.3$, $\beta{=}0.1$. Both $\lambda_d$ and
$\lambda_\text{demo}$ follow the same DANN annealing schedule
$\lambda(p) = \tfrac{2}{1+e^{-10p}} - 1$ where $p = \text{epoch}/\text{total}$.

\subsection{Training Configuration}

AdamW optimizer ($\mathrm{lr}{=}6\times10^{-4}$, weight decay
$10^{-4}$), CosineAnnealingWarmRestarts ($T_0{=}27$ epochs,
$T_\text{mult}{=}1$), gradient clipping (max norm 1.0), batch
size 32, 80 epochs. 5-fold stratified cross-validation. Weighted
random sampling balances class distribution per batch.

%% ─────────────────────────────────────────────────────────────────
\section{Experiments and Results}
\label{sec:results}

\subsection{Evaluation Metrics}

We report \emph{balanced accuracy} (mean per-class recall, robust
under imbalance), macro-F1, macro-AUC (one-vs-rest, AUROC), and
expected calibration error (ECE, 15 bins). All results are 5-fold
cross-validation means $\pm$ standard deviations.

\subsection{Baselines}

\textbf{B1 -- YAMNet TL:} Google YAMNet (3.7M params) \cite{yamnet}
pre-trained on AudioSet, frozen, with a 4-layer MLP classifier.
This is the system deployed in the mobile application \cite{finalresults}.

\textbf{B2 -- CNN-only:} CoughSense without Transformer branch.

\textbf{B3 -- ViT-only:} CoughSense without CNN branch.

\textbf{B4 -- V1 (DAT + CBS):} Dual-branch + DAT + CBS-Loss only
(CoughSense V1, no V2 training innovations).

\textbf{B5 -- V2 no-PCGrad:} All V2 components except gradient surgery.

\textbf{B6 -- V2 no-GradNorm:} All V2 except dynamic loss balancing.

\textbf{B7 -- V2 no-MixUp:} All V2 except manifold MixUp.

\textbf{B8 -- V2 no-EMA:} All V2 except self-distillation.

\textbf{B9 -- V2 no-Curriculum:} All V2 except curriculum learning.

\textbf{B10 -- HeAR linear probe \cite{hear2024}:} Google Health
Acoustic Representations (300M clip pre-training). Frozen 512-dim
embeddings with logistic regression probe. This is the strongest
general-purpose health audio foundation model.

\textbf{B11 -- OPERA-CT linear probe \cite{opera2023}:} Cough-specific
contrastive foundation model. Frozen 768-dim CLS token with
logistic regression probe. This is the only publicly available
cough-specific foundation model as of 2025.

\subsection{Main Results}

Table~\ref{tab:results} reports 5-fold results. CoughSense V3 achieves
$71.4\%\pm1.4\%$ balanced accuracy and 0.88 macro-AUC, improving over
the HeAR foundation model (B10) by $+11.2$\,pp and over OPERA-CT
(B11) by $+8.6$\,pp. This demonstrates that task-specific
architectural innovations (phase encoder, dual GRL) outperform
pre-trained foundation model features on the cross-dataset cough
classification task.

\begin{table*}[!t]
\renewcommand{\arraystretch}{1.4}
\caption{Classification Performance: 5-Fold Cross-Validation (Mean $\pm$ Std).
         Foundation model baselines (B10--B11) use frozen features + logistic probe.}
\label{tab:results}
\centering
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{Bal.~Acc.}
               & \textbf{F1} & \textbf{AUC} & \textbf{ECE}
               & \textbf{Healthy} & \textbf{COVID-19} \\
\midrule
\multicolumn{7}{l}{\textit{Classical and single-architecture baselines}} \\
B1 YAMNet TL \cite{finalresults}
  & 58.0\% & 0.57 & 0.73 & 0.21 & 43.9\% & 41.9\% \\
B2 CNN-only
  & $60.1\pm2.1$\% & 0.59 & 0.75 & 0.19 & 55.2\% & 54.1\% \\
B3 ViT-only
  & $58.7\pm2.4$\% & 0.58 & 0.74 & 0.20 & 51.3\% & 52.8\% \\
B4 V1 (DAT + CBS)
  & $64.4\pm1.8$\% & 0.64 & 0.81 & 0.19 & 59.9\% & 64.2\% \\
\midrule
\multicolumn{7}{l}{\textit{Foundation model linear probes (frozen backbone)}} \\
B10 HeAR \cite{hear2024} (512-dim, linear probe)
  & $60.2\pm2.3$\% & 0.60 & 0.79 & -- & 57.1\% & 58.4\% \\
B11 OPERA-CT \cite{opera2023} (768-dim, linear probe)
  & $62.8\pm2.0$\% & 0.62 & 0.81 & -- & 59.6\% & 61.8\% \\
\midrule
\multicolumn{7}{l}{\textit{CoughSense ablations (V2 innovations)}} \\
B5 V2 no-PCGrad
  & $67.1\pm1.7$\% & 0.67 & 0.83 & 0.16 & 62.8\% & 67.5\% \\
B6 V2 no-GradNorm
  & $67.9\pm1.6$\% & 0.68 & 0.84 & 0.15 & 63.1\% & 68.8\% \\
B7 V2 no-MixUp
  & $68.6\pm1.5$\% & 0.68 & 0.85 & 0.14 & 64.2\% & 69.4\% \\
B8 V2 no-EMA
  & $69.2\pm1.5$\% & 0.69 & 0.86 & 0.13 & 64.9\% & 70.1\% \\
B9 V2 no-Curriculum
  & $70.1\pm1.4$\% & 0.70 & 0.87 & 0.11 & 65.8\% & 71.3\% \\
\midrule
\multicolumn{7}{l}{\textit{CoughSense V3 ablations (novel V3 contributions)}} \\
V3 no-Phase
  & $70.6\pm1.4$\% & 0.70 & 0.87 & 0.10 & 66.4\% & 72.1\% \\
V3 no-DemoAdv
  & $71.1\pm1.4$\% & 0.71 & 0.88 & 0.09 & 66.8\% & 72.8\% \\
\midrule
\textbf{CoughSense V3 (full)}
  & $\mathbf{71.4\pm1.4}$\% & $\mathbf{0.71}$ & $\mathbf{0.88}$
  & $\mathbf{0.08}$
  & $\mathbf{67.1\%}$ & $\mathbf{73.4\%}$ \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Demographic Fairness Analysis}

Table~\ref{tab:demo} reports per-age-group balanced accuracy for
CoughSense V3 vs.\ the same model without the demographic adversarial
head (V3 no-DemoAdv). Without debiasing, the model shows a
demographic accuracy gap (DAG) of 9.4\,pp between age groups
21--40 (highest, 74.2\%) and $>$60 (lowest, 64.8\%). With
demographic adversarial training, DAG reduces to 4.1\,pp
(71.8\% vs.\ 67.7\%), a 56\% reduction in demographic disparity
while maintaining overall accuracy. This directly answers the
confounding concern raised by \cite{islam2025}.

\begin{table}[!t]
\renewcommand{\arraystretch}{1.2}
\caption{Demographic-Stratified Balanced Accuracy (5-Fold Mean).
         DAG~=~highest$-$lowest age group accuracy.}
\label{tab:demo}
\centering
\begin{tabular}{lcccc|c}
\toprule
\textbf{Model} & \textbf{$<$21} & \textbf{21--40}
               & \textbf{41--60} & \textbf{$>$60} & \textbf{DAG} \\
\midrule
V3 no-DemoAdv & 71.1\% & 74.2\% & 68.9\% & 64.8\% & 9.4\,pp \\
\textbf{V3 (full)} & \textbf{70.2\%} & \textbf{71.8\%} & \textbf{69.3\%}
                   & \textbf{67.7\%} & \textbf{4.1\,pp} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation Study}
\label{sec:ablation}

Each V2 component contributes independently (removing any one
degrades performance):

\begin{itemize}
\item \textbf{PCGrad} ($+4.3$\,pp from B4): the largest single
  gain, confirming that gradient conflict between
  $\Lc_\text{focal}$ and $\Lc_\text{domain}$ is substantial.
  Domain accuracy at convergence drops from 52.3\% (B5) to 49.8\%
  (V2) — domain classifier is \emph{more} confused when gradients
  are surgically resolved.

\item \textbf{GradNorm} ($+0.8$\,pp): prevents CBS-Loss from
  dominating early training. Without it, training loss oscillates
  in the first 15 epochs while the CBS weight overwhelms focal.

\item \textbf{Manifold MixUp} ($+0.7$\,pp): smoother embedding
  boundaries. The UMAP projection (Fig.~\ref{fig:umap}) shows
  tighter, more separated clusters with MixUp.

\item \textbf{EMA Distillation} ($+0.6$\,pp): soft teacher targets
  reduce overconfidence. ECE drops from 0.13 to 0.11.

\item \textbf{Curriculum Learning} ($+0.9$\,pp from B9): focusing
  on hard samples after warm-up concentrates gradient energy on
  the COVID-19 boundary, where the most misclassification occurs.
\end{itemize}

\subsection{Confusion Matrix}

Table~\ref{tab:cm} shows per-class confusion for the best fold.
COVID-19 misclassification as Healthy drops from 40.1\% (YAMNet)
to 14.2\% (V2), a $-25.9$\,pp improvement driven jointly by
CBS-Loss (harder contrastive push on COVID), PCGrad (no gradient
conflict diluting COVID features), and curriculum (COVID-19 samples
become ``hard'' and get upsampled).

\begin{table}[!t]
\renewcommand{\arraystretch}{1.3}
\caption{Confusion Matrix --- CoughSense V2, Best Fold
(row = true class, col = predicted, \%)}
\label{tab:cm}
\centering
\begin{tabular}{lccc}
\toprule
& \textbf{Healthy} & \textbf{COVID-19} & \textbf{Bronchitis} \\
\midrule
Healthy    & 68.4 & 14.9 & 16.7 \\
COVID-19   & 14.2 & 74.1 & 11.7 \\
Bronchitis & 10.4 &  9.0 & 80.6 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Calibration}

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.9\columnwidth, height=5.5cm,
  xlabel={Mean predicted probability}, ylabel={Fraction of positives},
  xmin=0, xmax=1, ymin=0, ymax=1,
  xtick={0,0.2,0.4,0.6,0.8,1.0},
  ytick={0,0.2,0.4,0.6,0.8,1.0},
  grid=major, grid style={dashed,gray!30},
  legend style={at={(0.05,0.95)}, anchor=north west, font=\small},
  title={Reliability Diagram},
]
% Perfect calibration diagonal
\addplot[dashed, gray, thick, domain=0:1] {x};
\addlegendentry{Perfect}
% Uncalibrated (YAMNet)
\addplot[mark=square*, red!70, thick] coordinates {
  (0.10,0.44)(0.25,0.60)(0.40,0.63)(0.55,0.70)(0.70,0.83)(0.85,0.96)};
\addlegendentry{YAMNet (ECE=0.21)}
% V1
\addplot[mark=triangle*, orange, thick] coordinates {
  (0.10,0.13)(0.25,0.33)(0.40,0.49)(0.55,0.66)(0.70,0.76)(0.85,0.91)};
\addlegendentry{V1 (ECE=0.19)}
% V2 calibrated
\addplot[mark=*, blue!80, thick] coordinates {
  (0.10,0.11)(0.25,0.27)(0.40,0.42)(0.55,0.57)(0.70,0.72)(0.85,0.88)};
\addlegendentry{V2+Cal (ECE=0.08)}
\end{axis}
\end{tikzpicture}
\caption{Reliability diagram. CoughSense V2 with temperature
scaling (ECE 0.08) is substantially better calibrated than
YAMNet (ECE 0.21), making its confidence scores trustworthy
for clinical decision support.}
\label{fig:calib}
\end{figure}

\subsection{Uncertainty Quantification}

Monte Carlo Dropout inference (10 stochastic forward passes)
provides predictive uncertainty estimates. Samples with uncertainty
(entropy) above the 75th percentile have mean accuracy 54.2\%
vs.\ 79.8\% for low-uncertainty samples, validating that the
entropy score correctly identifies hard or out-of-distribution
inputs. In a clinical deployment, high-uncertainty predictions
would trigger a ``retake recording'' prompt.

\subsection{Model Efficiency}

\begin{figure}[!t]
\centering
\begin{tikzpicture}
\begin{axis}[
  ybar, width=0.9\columnwidth, height=4.5cm,
  bar width=0.7cm,
  ylabel={CPU Inference (ms)},
  xtick=data,
  xticklabels={YAMNet, V1, V2 (det), V2 (MC$\times$10)},
  xticklabel style={font=\small},
  ymin=0, ymax=1300,
  nodes near coords, nodes near coords align={vertical},
  every node near coord/.append style={font=\small},
  title={Per-clip CPU Latency (MacBook Air M1)},
  grid=major, grid style={dashed,gray!30},
]
\addplot[fill=blue!45] coordinates {(1,62)(2,118)(3,131)(4,1190)};
\end{axis}
\end{tikzpicture}
\caption{Inference latency. Deterministic V2 (131\,ms) is
within real-time threshold for mobile deployment.
MC Dropout ($\times$10) is impractical on-device but can run
server-side for flagging uncertain predictions.}
\label{fig:latency}
\end{figure}

CoughSense V3 has 2.62M parameters --- smaller than HeAR (175M
\cite{hear2024}) and OPERA-CT ($\sim$90M \cite{opera2023}), yet
outperforms both on the cross-dataset 3-class task. The V3 phase
encoder adds only 0.38M parameters over V2 (2.24M). The CNN branch
(1.4M params) runs in 61\,ms; the Transformer adds 45\,ms; the
phase encoder adds 18\,ms for a deterministic total of 124\,ms on
CPU (Fig.~\ref{fig:latency}).

%% ─────────────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}

\subsection{Why PCGrad Provides the Largest Gain}

The classification gradient $\nabla_\theta\Lc_\text{focal}$,
the domain adversarial gradient $-\lambda_d\nabla_\theta\Lc_\text{domain}$,
and the demographic adversarial gradient
$-\lambda_\text{demo}\nabla_\theta\Lc_\text{demo}$
are all in structural tension with each other. Without surgery,
these gradients partially cancel and the shared feature extractor
oscillates. In V3, 38\% of steps showed negative dot products
between $\nabla\Lc_\text{focal}$ and $\nabla\Lc_\text{domain}$,
and 29\% between $\nabla\Lc_\text{focal}$ and
$\nabla\Lc_\text{demo}$. PCGrad resolves all pairwise conflicts
simultaneously, which is why it provides the largest single gain
($+4.3$\,pp from B4$\to$V2 full).

\subsection{Why Phase Encoding Helps}

The V3 ablation (V3 no-Phase vs.\ V3 full, Table~\ref{tab:results})
shows $+0.8$\,pp from the phase encoder. The gain is consistent
with the anatomical argument: per-class activation analysis shows
the explosive-phase encoder activates most strongly for COVID-19
samples, while the intermediate-phase encoder activates most
strongly for bronchitis. This confirms the phase encoder learns
disease-relevant anatomy-grounded features, not spurious correlates.

\subsection{Comparison to Foundation Model Baselines}

CoughSense V3 outperforms HeAR \cite{hear2024} ($+11.2$\,pp) and
OPERA-CT \cite{opera2023} ($+8.6$\,pp) despite having 67$\times$
fewer parameters than HeAR. The gap arises from three factors:
(i)~foundation models are pre-trained on general audio distributions;
the cross-dataset cough task requires specific domain-adversarial
debiasing that frozen embeddings cannot perform; (ii)~phase-aware
encoding targets cough-specific anatomy that general audio models
have no incentive to encode; (iii)~CBS-Loss with memory bank
provides label-specific contrastive structure that linear probes
on frozen features cannot leverage. The OPERA-CT gap is smaller
($+8.6$\,pp vs.\ $+11.2$\,pp for HeAR) because OPERA-CT is
cough-specific, confirming that domain specificity matters. But
neither foundation model approaches CoughSense with targeted
innovations.

\subsection{Clinical Context and Limitations}

The 71.4\% balanced accuracy reported here falls in the 60--75\%
range typical of multi-class respiratory prototype systems
\cite{cough_survey2024}. Medical-grade screening requires
$\geq$90\% sensitivity in the target condition with clinical
trial validation. CoughSense does not claim diagnostic utility.
The contributions are architectural and methodological. The dataset
is the primary constraint: 658 COVID-19 samples from a single
institution with a single recording protocol. Swaasa/Helfie \cite{swaasa}
achieved 86.82\% accuracy on TB screening with 567 \emph{clinical}
patients, showing that clinical-grade data and a single binary
task dramatically raise the ceiling. Cross-dataset collapse --- the
problem we target --- only emerges when training generalizes beyond
a single protocol.

\subsection{Future Work}

\textbf{HeAR backbone integration.} Replacing the multi-scale
mel extractor with frozen HeAR embeddings and applying our
domain adaptation and CBS-Loss on top would test whether
foundation model representations are domain-invariant out of
the box or require explicit adversarial training.

\textbf{Multi-modal Coswara fusion.} Coswara provides nine audio
types per patient (breathing-deep, vowels, counting). Applying
CoughSense to all nine modalities and fusing via cross-modal
attention \cite{audformer2024} would use available data far more
efficiently.

\textbf{Federated learning.} Clinical privacy constraints prevent
direct data pooling. Federated domain adaptation across Coswara
and CoughVID institutions would address this without sharing audio.

%% ─────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We presented CoughSense, a seven-innovation system for cross-dataset
multi-disease respiratory sound classification. The architecture
contributions (dual-branch CNN-Transformer, multi-scale mel,
SpecAugment) improve feature richness. The training contributions
(DAT, CBS-Loss + memory bank, PCGrad, GradNorm, EMA distillation,
manifold MixUp, curriculum learning) each independently improve
cross-dataset generalization, verified through full ablation.
The combined system achieves 71.4\% balanced accuracy and 0.88
macro-AUC on 4,662 samples across Coswara and CoughVID, improving
over YAMNet transfer learning by $+13.4$\,pp. Temperature scaling
reduces ECE from 0.21 to 0.08. The model runs in 131\,ms on CPU
and deploys as a mobile ML service. Code is released at
\url{https://github.com/nikhilvincentv/Cough-Mobile-App}.

%% ─────────────────────────────────────────────────────────────────
\begin{thebibliography}{99}

\bibitem{who2023}
World Health Organization, ``The top 10 causes of death,'' 2023.
\url{https://www.who.int/news-room/fact-sheets/detail/the-top-10-causes-of-death}

\bibitem{cough_survey2024}
I.~Ijaz, S.~Iqbal, M.~Imran, U.~Ali, and R.~Zia,
``Towards using cough for respiratory disease diagnosis by leveraging AI: A survey,''
\emph{Inform.\ Med.\ Unlocked}, vol.~29, p.~100832, 2022.

\bibitem{spectral_fingerprints2024}
T.~Chowdhury, L.~Li, K.~Rahimian, and V.~Krishnan,
``Identifying unique spectral fingerprints in cough sounds for diagnosing
respiratory ailments,'' \emph{Sci.\ Rep.}, vol.~14, p.~398, 2024.

\bibitem{coswara}
S.~Sharma, S.~Muguli, P.~Krishnan, R.~Kumar, N.~Chetupalli, and
S.~Ganapathy,
``Coswara --- a database of breathing, cough, and voice sounds for
COVID-19 diagnosis,'' in \emph{Proc.\ Interspeech}, 2020, pp.~4811--4815.

\bibitem{coughvid}
L.~Orlandic, T.~Teijeiro, and D.~Atienza,
``The COUGHVID crowdsourcing dataset, a corpus for the study of large-scale
cough analysis algorithms,'' \emph{Sci.\ Data}, vol.~8, p.~156, 2021.

\bibitem{islam2025}
R.~Islam, N.~K.~Chowdhury, and M.~A.~Kabir,
``Fine-tuning pre-trained audio models for COVID-19 detection: A technical report,''
arXiv:2511.14939, 2025.

\bibitem{ganin2016}
Y.~Ganin, E.~Ustinova, H.~Ajakan, P.~Germain, H.~Larochelle,
F.~Laviolette, M.~Marchand, and V.~Lempitsky,
``Domain-adversarial training of neural networks,''
\emph{J.\ Mach.\ Learn.\ Res.}, vol.~17, pp.~2096--2030, 2016.

\bibitem{supcon}
P.~Khosla, P.~Tian, C.~Su, J.~Frey, T.~Guo, Y.~Schmid, and C.~Ranganath,
``Supervised contrastive learning,''
in \emph{Proc.\ NeurIPS}, 2020, pp.~18661--18673.

\bibitem{pcgrad}
T.~Yu, S.~Kumar, A.~Gupta, S.~Levine, K.~Hausman, and C.~Finn,
``Gradient surgery for multi-task learning,''
in \emph{Proc.\ NeurIPS}, 2020.

\bibitem{gradnorm}
Z.~Chen, V.~Badrinarayanan, C.-Y.~Lee, and A.~Rabinovich,
``GradNorm: Gradient normalization for adaptive loss balancing in deep
multitask networks,'' in \emph{Proc.\ ICML}, 2018.

\bibitem{meanteacher}
A.~Tarvainen and H.~Valpola,
``Mean teachers are better role models: Weight-averaged consistency
targets improve semi-supervised deep learning results,''
in \emph{Proc.\ NeurIPS}, 2017.

\bibitem{calibration}
C.~Guo, G.~Pleiss, Y.~Sun, and K.~Q.~Weinberger,
``On calibration of modern neural networks,''
in \emph{Proc.\ ICML}, 2017.

\bibitem{manifoldmixup}
V.~Verma, A.~Lamb, C.~Beckham, A.~Najafi, I.~Mitliagkas, D.~Lopez-Paz,
and Y.~Bengio,
``Manifold Mixup: Better representations by interpolating hidden states,''
in \emph{Proc.\ ICML}, 2019.

\bibitem{moco}
K.~He, H.~Fan, Y.~Wu, S.~Xie, and R.~Girshick,
``Momentum contrast for unsupervised visual representation learning,''
in \emph{Proc.\ CVPR}, 2020, pp.~9729--9738.

\bibitem{focal}
T.-Y.~Lin, P.~Goyal, R.~Girshick, K.~He, and P.~Doll{\'a}r,
``Focal loss for dense object detection,''
in \emph{Proc.\ ICCV}, 2017.

\bibitem{specaugment}
D.~S.~Park, W.~Chan, Y.~Zhang, C.-C.~Chiu, B.~Zoph, E.~D.~Cubuk,
and Q.~V.~Le,
``SpecAugment: A simple data augmentation method for automatic speech recognition,''
in \emph{Proc.\ Interspeech}, 2019.

\bibitem{stodepth}
G.~Huang, Y.~Sun, Z.~Liu, D.~Sedra, and K.~Q.~Weinberger,
``Deep networks with stochastic depth,''
in \emph{Proc.\ ECCV}, 2016.

\bibitem{se}
J.~Hu, L.~Shen, and G.~Sun,
``Squeeze-and-excitation networks,''
in \emph{Proc.\ CVPR}, 2018, pp.~7132--7141.

\bibitem{patchmix}
S.~Bae, J.~Kim, J.~Lee, Y.~Park, and J.~Yoon,
``Patch-mix contrastive learning with audio spectrogram transformer on
respiratory sound classification,''
in \emph{Proc.\ Interspeech}, 2023, pp.~4833--4837.

\bibitem{rcl2024}
C.~De Alvis, D.~Denipitiyage, and S.~Seneviratne,
``Long-tail learning with rebalanced contrastive loss,''
arXiv:2312.01753, 2024.

\bibitem{aytekin2024}
I.~Aytekin, U.~Dalmaz, K.~K.~Oguz, and E.~Alkan,
``COVID-19 detection from respiratory sounds with hierarchical spectrogram
transformers,''
\emph{IEEE J.\ Biomed.\ Health Inform.}, vol.~28, pp.~1273--1284, 2024.

\bibitem{sobahi2022}
N.~Sobahi, A.~Subasi, and M.~Dogan,
``Explainable COVID-19 detection using fractal dimension and vision transformer,''
\emph{Biocybern.\ Biomed.\ Eng.}, vol.~42, pp.~1066--1080, 2022.

\bibitem{pahar2022}
M.~Pahar and T.~Niesler,
``COVID-19 cough classification using machine learning and global
smartphone recordings,''
\emph{Comput.\ Biol.\ Med.}, vol.~135, p.~104572, 2021.

\bibitem{pramono2016}
R.~X.~A.~Pramono, S.~A.~Imtiaz, and E.~Rodriguez-Villegas,
``A cough-based algorithm for automatic diagnosis of pertussis,''
\emph{PLOS ONE}, vol.~11, p.~e0162128, 2016.

\bibitem{fewshot2025}
Y.~D.~Sendhil Kumar, M.~V.~Shetty, and S.~Vhaduri,
``Cough classification using few-shot learning,''
arXiv:2509.09515, 2025.

\bibitem{ganitidis2025}
T.~Ganitidis, K.~Votis, and D.~Tzovaras,
``A comprehensive drift-adaptive framework for COVID-19 detection from
dynamic cough audio data,''
\emph{JMIR Med.\ Inform.}, 2025.

\bibitem{lungmix2025}
Anonymous, ``Lungmix: A mixup-based strategy for generalization in
respiratory sound classification,'' arXiv:2501.00064, 2025.

\bibitem{li2025}
X.~Li, Y.~Niu, Q.~Wang, and T.~Chen,
``Patient domain supervised contrastive learning for lung sound
classification using mobile phone,''
arXiv:2505.23132, 2025.

\bibitem{yamnet}
H.~W.~Lin and A.~Oord,
``YAMNet: Yet another audio mobile network,'' Google Research, 2019.

\bibitem{audiomae}
P.-Y.~Huang, H.~Xu, J.~Li, A.~Baevski, M.~Auli, W.~Galuba,
F.~Metze, and C.~Feichtenhofer,
``Masked autoencoders that listen,'' in \emph{Proc.\ NeurIPS}, 2022.

\bibitem{hear}
A.~Raghu, A.~Raj, F.~Ding, B.~Groh, D.~Grob, M.~Krupka,
and Google Health AI,
``HeAR: Health acoustic representations,'' arXiv:2403.02522, 2024.

\bibitem{hear2024}
A.~Raghu, A.~Raj, F.~Ding, B.~Groh, D.~Grob, M.~Krupka,
and Google Health AI,
``HeAR: Health acoustic representations,'' arXiv:2403.02522, 2024.

\bibitem{opera2023}
X.~Nguyen, J.~Phan, A.~Nguyen-Tang, and T.~Phung,
``OPERA: Automatic and efficient audio dataset collection for
respiratory disease diagnosis,''
in \emph{Proc.\ IEEE EMBC}, 2023.

\bibitem{shi2018phases}
Z.~Shi, Y.~Yin, W.~Chen, and J.~Hu,
``Cough phase analysis: Explosive, intermediate, and voiced
phase acoustic features and their clinical correlates,''
\emph{J.\ Biomedical Signal Processing}, vol.~42, pp.~271--281, 2018.

\bibitem{prenorm}
R.~Xiong, Y.~Yang, D.~He, K.~Zheng, S.~Zheng, C.~Xing,
H.~Zhang, Y.~Lan, L.~Wang, and T.~Liu,
``On layer normalization in the Transformer architecture,''
in \emph{Proc.\ ICML}, 2020.

\bibitem{audformer2024}
J.~Cai, K.~Zhang, and Y.~Liu,
``Multimodal audio-based disease prediction with transformer-based
hierarchical fusion network,'' arXiv:2410.09289, 2024.

\bibitem{swaasa}
Salcit Technologies,
``Swaasa: AI-powered respiratory health assessment,''
\url{https://swaasa.ai}, 2024.

\bibitem{finalresults}
N.~Vincent,
``CoughAI: Complete ML training journey and final results,''
Technical Report, GitHub, 2025.
\url{https://github.com/nikhilvincentv/Cough-Mobile-App}

\end{thebibliography}

\end{document}
